---
title: "Homework 2"
author: "Paul J Anderson"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r clear-workspace, include=FALSE}
# clear all objects from the environment
# rm(list = ls())

# clear the console — only works in RStudio, not R Markdown output
# cat("\014")

# this can also be done with the environment pane > broom icon
```

```{r setup, include=FALSE}
# do not include coding chunks in knit
knitr::opts_chunk$set(echo = TRUE)
```

```{r load in libraries}
library(vcdExtra)
library(Sleuth3)
library(tidyverse)
library(gnm)
# detach("package:epitools", unload = TRUE, character.only = TRUE)
```

What to submit (gradescope):
1. .Rmd
2. pdf (include code in the R question)

## R Questions: 

### Q1. (4 points) It is often argued that victims of violence exhibit more violent behavior toward others. To study this hypothesis, a researcher searched court records to find 908 individuals who had been victims of abuse as children. She then found 667 individuals, with similar demographic characteristics, who had not been abused as children. Based on a search through subsequent years of court records, she was able to determine how many in each of these groups became involved in violent crimes, as shown in the following table.

```{r}
# recreate the given 2 x 2 table
crime_table <- matrix(
  c(102, 806, 53, 614),  # fill by column
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    "Group" = c("Abuse Victim", "Control"),
    "Violent Crime" = c("Yes", "No")
  )
)

# simple print
crime_table

# output for knit
knitr::kable(as.data.frame(crime_table), caption = "Involvement in Violent Crime by Group")
```

### The researcher concluded: ”Early childhood victimization has demonstrable long-term consequences for violent criminal behavior.” Conduct your own analysis of the data and comment on this conclusion. Is there evidence of a difference between the two groups? Is the strength of the causal implication of this statement justified by the data from this study?

ANSWER: Since both variables are categorical, from a dataset of sufficient size (>5), we are able to use the Pearson's chi-squared test for independence with Yates' continuity correction (by default) to test the H_0 that the distribution of these two proportions are independent, we use 1 DF for this 2 x 2 contingency table.

We find that there is moderate evidence (p-value = 0.03765) to support the rejection of the null hypothesis that early childhood victimization and long-term violent criminal behavior are independent. While these data show a link between these two variables, this test is observational, not experimental. Therefore, the strong causal implication made in the researchers statement is not justified by the data alone. Additional controls or experimental designs would be needed to reach this level of causal inference.

```{r}
chisq.test(crime_table)
```

## Conceptual Questions

### Q2. (1 point) During an investigation of the U.S. space shuttle Challenger disaster, it was learned that project managers had judged the probability of mission failure to be 0.00001, whereas engineers working on the project had estimated failure probability at 0.005. The difference between these two probabilities, 0.00499, was discounted as being too small to worry about. Is a different picture provided by considering odds? How is that interpreted?

ANSWER: Odds can highlight risk more intuitively than raw probabilities. We know that odds ratios are calculated in the following way from our readings and lecture, where p is the probability of failure in this context:

$$
\omega = \frac{p}{1-p}
$$
By calculating the two odds ratios and comparing them as a proportion, we get the increased likelihood of failure from one estimate to another. In this case, using odds ratios show the engineering estimate of failure to by over 500 times (over 50,000%!) that of the project managers and could have easily been identified as critical and corrective action could have been implemented.

```{r}
(odds_proj <- 0.00001/(1-0.00001))
(odds_eng <- 0.005/(1-0.005))

(proportion_proj_eng <- odds_eng/odds_proj)

(percent_prop_proj_eng <- proportion_proj_eng*100)
```

### Q3. (2 points) Suppose that 90% of orange tabby cats are male. Determine if the following statements are true or false, and explain your reasoning.

    (a) The distribution of sample proportions of random samples of size 30 is left skewed.
    
ANSWER: FALSE. Sample proportions cant exceed 1, so this is a boundary. The sample size is relatively small and with 90% of taby's as males, the data would be clustered around this upper boundary. So, we would say that the distribution of the sample proportions of random samples of size 30 is right skewed, not left skewed.
    
    (b) Using a sample size that is 4 times as large will reduce the standard error of the sample proportion by one-half.
    
ANSWER: TRUE. SE$\hat{p}$ is calculated by the formula below. If we multiple n by 4, we can factor this out as \sqrt{1/4} which further reduces to 1/2. 1/2 times $\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$ = 1/2 times SE$\hat{p}$.
$$
SE\hat{p} = \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}
$$
    
    (c) The distribution of sample proportions of random samples of size 140 is approximately normal.

ANSWER: TRUE. We must assess whether the conditions of normality are met. Based on our calculations, both the number of successes (np) and the number of failures (n(1-p)) are >= 10. Therefore, the sample proportions of random samples of size 140 for the given proportion are approximately normal.
```{r}
# define the sample size and population proportion
n <- 140
p <- 0.9

# calculate the number of successes
np <- n * p
# calculate the number of failures
n1_minus_p <- n * (1 - p)

# check if there are enough successes (np >= 10)
if (np >= 10) {
  print("There are enough successes.")
} else {
  print("There are not enough successes.")
}

# check if there are enough failures (n(1 - p) >= 10)
if (n1_minus_p >= 10) {
  print("There are enough failures.")
} else {
  print("There are not enough failures.")
}
```

    (d) The distribution of sample proportions of random samples of size 280 is approximately normal.

ANSWER: TRUE. We repeate the calculations for success and failure thresholds. Both return values >= 10. Therefore, the sample proportions of random samples of size 280 for the given proportion are approximately normal.

```{r}
# define the sample size and population proportion
n <- 280
p <- 0.9

# calculate the number of successes
np <- n * p
# calculate the number of failures
n1_minus_p <- n * (1 - p)

# check if there are enough successes (np >= 10)
if (np >= 10) {
  print("There are enough successes.")
} else {
  print("There are not enough successes.")
}

# check if there are enough failures (n(1 - p) >= 10)
if (n1_minus_p >= 10) {
  print("There are enough failures.")
} else {
  print("There are not enough failures.")
}
```

### Q4. (2 points) A 2010 survey asked 827 randomly sampled registered voters in California, ”Do you support or do you oppose drilling for oil and natural gas off the coast of California? Or do you not know enough to say?” Below is the distribution of responses, separated based on whether or not the respondent graduated from college.

ANSWER: Here we recreate the plot given in the module HW and excluded the totals row because it didn't provide any useful information we can't calculate easily.

```{r}
# recreate the given 3 x 2 table - we excluded the totals row
drill_table <- matrix(
  c(154, 180, 104, 132, 126, 131),  # fill by column
  nrow = 3,
  byrow = FALSE,
  dimnames = list(
    "Position" = c("Support", "Oppose", "Do not know"),
    "College Grad" = c("Yes", "No")
  )
)

# simple print
drill_table

# output for knit
knitr::kable(as.data.frame(drill_table), caption = "Involvement in Violent Crime by Group")
```

    (a) What percents of college graduates and non-college graduates in this sample do not know enough to have an opinion on drilling for oil and natural gas off the coast of California (i.e.,report two percent values)?
    
ANSWER:
% college grads who do not know: 23.7%
% non-college grads who do not know: 33.7%

```{r}
# % college grads who do not know:
104/438*100
# % non-college grads who do not know:
131/389*100
```

    (b) Conduct a hypothesis test to determine whether there is evidence that the proportion of college graduates who do not have an opinion on this issue is different from that of non-college graduates.
    
ANSWER: 
This is a frequency table of a 3x2 matrix with categorical variables. The dataset is sufficiently large for performing a chi-squared test. The null hypothesis is that the proportions of college graduates and non-college graduates who do not have an opinion on the issue are equal (the two variables are independent).

Using the chi-squared test for independence, we obtain a p-value of 0.003246, which provides convincing evidence to reject the null hypothesis. This suggests that the proportions of college graduates and non-college graduates who do not have an opinion on drilling are significantly different from each other. In other words, education level and opinion on the issue are dependent.

Thus, the data support the conclusion that for California registered voters in 2010, education level is associated with the likelihood of not having an opinion on drilling for oil and natural gas off the coast of California.

```{r}
chisq.test(drill_table)
```

### Q5. (1 point) A study of British male physicians noted that the proportion who died from lung cancer was 0.0140 per year for cigarette smokers and 0.00010 per year for nonsmokers. Additionally, the proportion who died from heart disease was 0.00669 for smokers and 0.00413 for nonsmokers. Which response (lung cancer or heart disease) is more strongly related to cigarette smoking, in terms of the reduction in deaths that could occur with the absence of smoking?

ANSWER: Based on the odds ratios, lung cancer is approximately 87.4 times more strongly related to cigarette smoking than heart disease. This means that smoking has a far greater effect on the risk of lung cancer than on heart disease, and therefore, eliminating smoking would likely have a much greater impact on reducing lung cancer deaths.

```{r}
lung_smokers <- 0.0140
lung_nonsmokers <- 0.00010

(omega_lung = (lung_smokers/(1-lung_smokers))/(lung_nonsmokers/(1-lung_nonsmokers)))

heart_smokers <- 0.00669
heart_nonsmokers <- 0.00413

(omega_heart = (heart_smokers/(1-heart_smokers))/(heart_nonsmokers/(1-heart_nonsmokers)))

# calculate the proportion
omega_lung/omega_heart
```

